---
title: "EDS 223 Week 6 Lab"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Load packages

```{r}
library(sf) # vector handling
library(terra) # raster handling
library(tidyverse)
library(tmap) # map making
library(spData) # spatial data
library(spDataLarge) # spatial data
```

## Load data

```{r}
# load Zion park boundary (vector object to rasterize)
boundary <- read_sf(system.file("vector/zion.gpkg", package = "spDataLarge")) 

# load elevation raster to use as template raster object
elevation <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))

# check and transform coordinate reference systems
if(crs(elevation) == crs(boundary)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  boundary <- st_transform(boundary, st_crs(elevation))
}
```

#### Check geometry type of boundary 

```{r}
if(sf::st_geometry_type(boundary) == "POLYGON") {
  print("polygon data") 
} else{
  print("not polygon data")
}
```

#### Make the park boundary a line vector class

```{r}
# update park boundary object name to clarify that it's a polygon
park_polygon <- boundary

# create line version park boundary
park_border <- boundary %>%
  sf::st_cast(., "MULTILINESTRING") # '.' --> use the previous variable 


# make a map to compare these datasets 
map1 <- tm_shape(park_polygon) + 
  tm_polygons() + 
  tm_layout(main.title = "polygon")

map2 <- tm_shape(park_border) + 
  tm_lines() + 
  tm_layout(main.title = "lines")

tmap_arrange(map1, map2, nrow = 1)
```

## Rasterize

```{r}
# rasterize the park boundary layers
polygon_raster <- terra::rasterize(park_polygon, # layer to rasterize
                                   elevation) # template raster 
border_raster <- terra::rasterize(park_border, 
                                  elevation)
                                   

map3 <- tm_shape(polygon_raster) + 
  tm_raster() + 
  tm_layout(main.title = "polygon raster", 
            legend.outside = TRUE)

map4 <- tm_shape(border_raster) + 
  tm_raster() + 
  tm_layout(main.title = "border raster", 
            legend.outside = TRUE)

tmap_arrange(map3, map4, nrow = 1)
```

## Compare raster resolutions

```{r}
# create low spatial resolution template raster from scratch 
template_raster_low <- terra::rast(ext(park_polygon), resolution = 0.05,
                        crs = st_crs(park_polygon))

# create high spatial resolution template raster from scratch 
template_raster_high <- terra::rast(ext(park_polygon), resolution = 0.01,
                        crs = st_crs(park_polygon))

# rasterize to low resolution template raster
polygon_raster_low <- terra::rasterize(park_polygon, 
                                       template_raster_low)

# rasterize to high resolution template raster
polygon_raster_high <- terra::rasterize(park_polygon, 
                                        template_raster_high)



# plot 
map1 <- tm_shape(polygon_raster_low) +
  tm_raster() +
  tm_layout(main.title = "low resolution")

map2 <- tm_shape(polygon_raster_high) +
  tm_raster() +
  tm_layout(main.title = "high resolution")

tmap_arrange(map1, map2, nrow = 1)
```

## Rasterizing point data 

```{r}
# define point data
cycle_hire_osm <- spData::cycle_hire_osm

# transform to projected CRS
cycle_hire_osm_projected = sf::st_transform(cycle_hire_osm, "EPSG:27700")

# define raster template
template_raster <- terra::rast(ext(cycle_hire_osm_projected), resolution = 1000,
                       crs = crs(cycle_hire_osm_projected))
```

In the following code chunk we’ll perform three versions of rasterization to produce the following:

1. raster representing the presence/absence of cycle hire points
2. raster representing the number of cycle hire points
3. raster representing the total capacity of cycle hire points

```{r}
# presence/absence of cycle hire points 
ch_raster1 <- terra::rasterize(cycle_hire_osm_projected, template_raster)

# number of cycle hire points 
ch_raster2 <- rasterize(cycle_hire_osm_projected, template_raster, 
                       fun = "length")

# total capacity of cycle hire points
ch_raster3 <- rasterize(cycle_hire_osm_projected, template_raster, 
                       field = "capacity", fun = sum, na.rm = TRUE)



# plotting
map1 <- tm_shape(cycle_hire_osm_projected) +
  tm_symbols(col = "capacity") +
  tm_layout(main.title = "original points")

map2 <- tm_shape(ch_raster1) +
  tm_raster(title = "presence") +
  tm_layout(main.title = "presence/absence")

map3 <- tm_shape(ch_raster2) +
  tm_raster(title = "Hire points (n)") +
  tm_layout(main.title = "count of points")

map4 <- tm_shape(ch_raster3) +
  tm_raster(title = "Capacity (n bikes)") +
  tm_layout(main.title = "sum of capacity")

tmap_arrange(map1, map2, map3, map4, nrow = 2)
```
























